---
title: "Musical scales"
output: 
  html_document: 
    toc: yes
    toc_float: true
---


```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
# Generate this notebook to a pdf document
# cd ~/rp/paulrougieux.github.io && Rscript -e "rmarkdown::render('musical_scales.Rmd', 'pdf_document')"
library(dplyr)
library(tidyr)
```

# Introduction

The purpose of this notebook is to display notes used in various music
scales.


# Setup

Create a data frame that contains the twelve notes of the chromatic scale used later in the analysis.

```{r}
notes <- tribble(
    ~it,   ~en, 
    "do",  "C",
    "do#", "C#" ,
    "re",  "D",
    "mib", "Eb",
    "mi",  "E",
    "fa",  "F",
    "fa#", "F#",
    "sol", "G",
    "sol#","G#",
    "la",  "A",
    "sib", "Bb",
    "si",  "B",
)
notes$n = as.integer(rownames(notes))

# Duplicate the notes so that increments of scale number below don't fall out of the range.
notes2 <- rbind(notes, notes)

# A function to generate the line of a markdown table separated by pipes "|"
display <- function(x){cat("", x, "\n", sep="|")}

# The scale corresponding to the white notes of the piano
major_heptatonic <- c(0, 2, 4, 5, 7, 9, 11)
notes[major_heptatonic + 1,]

# A function to display a table with the given scale shifted to all the note of
# the major heptatonic scale
display_scale <- function(input_scale){
    display(input_scale)
    display(rep("---", length(input_scale)))
    for (i in major_heptatonic + 1){
        display(notes2$it[input_scale + i])
    }

    for (i in major_heptatonic + 1){
        display(notes2$en[input_scale + i])
    }
}
```

# Heptatonic scale

[Heptatonic scale](https://en.wikipedia.org/wiki/Heptatonic_scale)

> "A heptatonic scale is a musical scale that has seven pitches (a.k.a. tones)
> per octave. "

## Heptatonic major scale

```{r,  results='asis'}
display_scale(major_heptatonic)
```

The Bb or Sib scale
```{r}
notes2$it[major_heptatonic+11]
```

## Heptatonic minor scale

```{r,  results='asis'}
display_scale(c(0, 2, 3, 5, 7, 8, 10))
```

Note how the notes of the C major scale are the same as the notes of the A
minor scale. The notes of the F major scale are the same as the notes of the D
minor scale. This is explained in [minor
scale](https://en.wikipedia.org/wiki/Minor_scale) and the correspondance
between minor and major scales is visible on the [circle of
fifths](https://en.wikipedia.org/wiki/Circle_of_fifths).


## Chords in the scale of

For each scale, show the major, minor and 7 chords that are in that scale.

```{r message=FALSE, warning=FALSE, results="asis"}
major <- c(0, 4, 7)
minor <- c(0, 3, 7)
major7 <- c(0, 4, 7, 10)
notes2$it[major + 1]
notes2$it[minor + 1]
major_chords <- notes
major_chords$m4 = slice(notes2,(4+1):(4+12))
major_chords$m7 = slice(notes2,(7+1):(7+12))
minor_chords <- notes
minor_chords$m3 = slice(notes2,(3+1):(3+12))
minor_chords$m7 = slice(notes2,(7+1):(7+12))
major7_chords <- notes
major7_chords$m4 = slice(notes2,(4+1):(4+12))
major7_chords$m7 = slice(notes2,(7+1):(7+12))
major7_chords$m10 = slice(notes2,(10+1):(10+12))


for (i in 1:nrow(notes)){
    cat("\nThe scale: ")
    cat(notes2$it[major_heptatonic + i])
    major_match = c()
    minor_match = c()
    major7_match = c()
    for (j in 1:nrow(notes)){
        if (length(setdiff(notes2$n[major + j], notes2$n[major_heptatonic + i])) == 0L){
            major_match = c(major_match, j)
        }
        if (length(setdiff(notes2$n[minor + j], notes2$n[major_heptatonic + i])) == 0L){
            minor_match = c(minor_match, j)
        }
        if (length(setdiff(notes2$n[major7 + j], notes2$n[major_heptatonic + i])) == 0L){
            major7_match = c(major7_match, j)
        }
    }
    cat(" contains the major chords ")
    major_slice <- slice(major_chords,major_match)
    cat(major_slice$it)
    cat(" the minor chords ")
    minor_slice <- slice(minor_chords,minor_match)
    cat(paste0(minor_slice$it, "(m)"))
    cat(" and the major7 chord ")
    major7_slice = slice(major7_chords,major7_match)
    cat(paste0(major7_slice$it, "7"))
    chords <- unnest(major_slice, cols=c(m4,m7), names_repair="universal") %>%
        bind_rows(unnest(minor_slice, cols=c(m3,m7), names_repair="universal")) %>%
        bind_rows(unnest(major7_slice, cols=c(m4,m7,m10), names_repair="universal"))
    names(chords) <- paste0(rep(c("it", "en","n"),4),rep(1:4,each=3))
    options(knitr.kable.NA = '')
    chords %>% knitr::kable() %>% print()
}

```


# Pentatonic scales

## Major pentatonic

[Major pentatonic scale](https://en.wikipedia.org/wiki/Pentatonic_scale#Major_pentatonic_scale)

> "The major pentatonic scale may be thought of as a gapped or incomplete major
> scale.[8] However, the pentatonic scale has a unique character and is
> complete in terms of tonality. One construction takes five consecutive
> pitches from the circle of fifths;[9] starting on C, these are C, G, D, A,
> and E. Transposing the pitches to fit into one octave rearranges the pitches
> into the major pentatonic scale: C, D, E, G, A."


```{r,  results='asis'}
major_pentatonic <- c(0, 2, 4, 7, 9)
display_scale(major_pentatonic)
```

## Minor pentatonic

[Minor pentatonic
scale](https://en.wikipedia.org/wiki/Pentatonic_scale#Minor_pentatonic_scale)

> " It may also be considered a gapped blues scale. The C minor pentatonic
> is C, E-flat, F, G, B-flat. The A minor pentatonic, the relative minor of C
> pentatonic, comprises the same tones as the C major pentatonic, starting on
> A, giving A, C, D, E, G."

```{r,  results='asis'}
display_scale(c(0, 3, 5, 7, 10))
```


# Hexatonic scales

[Blues scale Hexatonic](https://en.wikipedia.org/wiki/Blues_scale#Hexatonic)

> "The hexatonic, or six-note, blues scale consists of the minor pentatonic
> scale plus the b5th degree of the original heptatonic scale." 

```{r,  results='asis'}
hexatonic <- c(0, 3, 5, 6, 7, 10)
display_scale(hexatonic)
```

