---
title: "R Commands"
author: "Paul Rougieux"
date: "17 August 2017"
output: 
  html_document: 
    toc: yes
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```


This page is the continuation of [my blog post on R commands](https://paulremote.blogspot.de/2014/04/r-commands.html). 
On the blog, see also [why use R](http://paulremote.blogspot.fr/2013/10/why-r.html) and the [RSS feed of posts labelled R](http://paulremote.blogspot.com/feeds/posts/default/-/R).

See also documentation at:

 * The Comprehensive R Archive Network [manuals](https://cran.r-project.org/manuals.html)
 * Tidyverse [packages](http://www.tidyverse.org/packages/)
 


# Set operations

```{r}
x = letters[1:3]
y = letters[3:5]
union(x, y)
intersect(x, y)
setdiff(x, y)
setdiff(y, x)
setequal(x, y)
```



# Information about your R system
```{r eval=FALSE}
sessionInfo()
installed.packages()
```


# Handling files

    getwd()
    list.files(tempdir()) 
    dir.create("blabla")
    read.csv("data.csv")

# Vectors

[SO Answer concerning indexing up to end of vector/matrix](https://stackoverflow.com/a/7500961/2641825) "Sometimes it's easier to tell R what you don't want"

    x <- c(5,5,4,3,2,1)
    x[-(1:3)]
    x[-c(1,3,6)]

# Lists
Given a list structure x, unlist simplifies it to produce a vector which contains all the atomic components which occur in x.
```{r}
l1 <- list(a="a", b="2,", c="pi+2i")
str(l1)
unlist(l1) # a character vector 
str(unlist(l1))
```


# Strings

```{r}
message("Using the following letters: ", paste(letters, collapse=","), ".")
```

# S3 methods
x<-1



List all available methods for a class:

    methods(class="lm") 


# One liners
Remove all objects in the workspace except one :

    rm(list=ls()[!ls()=="object_to_keep"]) 
    rm(list=ls()[!ls()=="con"]) # Remove all except a database connection


# knitr
## kable to create tables
```{r }
cat(kable(head(iris, 1), format = "html"))
cat(kable(head(iris, 1), format = "latex"))
cat(kable(head(iris, 1), format = "markdown"))
```

## Setting knitr options
Those 2 commands are different.
Sets the options for chunk, within a knitr chunk inside the .Rmd document

    opts_chunk$set(fig.width=10)

 Sets the options for knitr outside the .Rmd document

    opts_knit$set()

## R markdown python engine

Rstudio: [R Mardown python engine](https://rstudio.github.io/reticulate/articles/r_markdown.html)

# Tidyverse
## dplyr
pipes

```{r warning=FALSE, message=FALSE}
library(dplyr)
cars %>%
    group_by(speed) %>%
    print() %>% # works because the print function returns its argument
    summarise(numberofcars = n(),
              min = min(dist),
              mean = mean(dist),
              max = max(dist)) 
```



group_by() creates a tbl_df objects which is a wrapper around a data.frame to enable some functionalities. Note that print returns its output on a tbl_df object. So print() can be used inside the pipe without stopping the workflow.


### Mutate

Mutate multiple variables in the dataframe at once using the `vars()` helper function to scope the mutation:
```{r} 
iris %>%
    mutate_at(vars(starts_with("Petal")), round) %>%
    head()
```


## tidyr

[tidyr vignette on tidy data](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html) 
In the section on "Multiple types in one table":

> Datasets often involve values collected at multiple levels, on different types of observational units. During tidying, each type of observational unit should be stored in its own table. This is closely related to the idea of database normalisation, where each fact is expressed in only one place. It’s important because otherwise inconsistencies can arise.

> Normalisation is useful for tidying and eliminating inconsistencies. However, there are few data analysis tools that work directly with relational data, so analysis usually also requires denormalisation or the merging the datasets back into one table.

Example use of `tidyr::nest()` to generate a group of plots: 
[make ggplot2 purrr](http://www.brodrigues.co/blog/2017-03-29-make-ggplot2-purrr/).

```{r eval=FALSE}
library(tidyr)
library(dplyr)
library(purrr)
library(ggplot2)
piris <- iris %>%  
    group_by(Species) %>% 
    nest() %>% 
    mutate(plot = map2(data, Species, 
                       ~ggplot(data = .x, 
                               aes(x = Petal.Length, y = Petal.Width)) + 
                           geom_point() + ggtitle(.y)))
piris$plot[1]
piris$plot[3]
piris$plot[2]
```

# Discussions

* Gavin Simpson [My aversion to pipers](http://www.fromthebottomoftheheap.net/2015/06/03/my-aversion-to-pipes/) shows an Hadley tweet explaining that pipe might not be good in package development.


# Teaching R

* [Teach the tidyverse to beginners](http://varianceexplained.org/r/teach-tidyverse/)

* [Vectorization in R, Why?](http://www.noamross.net/blog/2014/4/16/vectorization-in-r--why.html)

* [Recommendations for resources for after workshops](https://github.com/carpentries/conversations/issues/18)


# Plotting with ggplot2

    geom_bar
    geom_tile + a gradient produce heat maps


# Creating a package
You might want to read the CRAN manual on [Writing R Extensions](https://cran.r-project.org/doc/manuals/R-exts.html), and its section on [Package dependencies](https://cran.r-project.org/doc/manuals/R-exts.html#Package-Dependencies).
See also Hadley's book on R package and its section on [Namespace](http://r-pkgs.had.co.nz/namespace.html)

Use the devtools library to start a package folder structure:

    devtools::create("package_name")

Use git to track code modifications (shell commands):

    $ cd package_name
    $ git init

## Documentation

The roxygen2 package helps with function documentation.
Documentation can be written in the form of comments #'
tags such as `@param` and `@description` structure the documentation of each function.

For an introcution to roxygen2, call
`vignette("roxygen2", package = "roxygen2")` at the R prompt.

Since roxygen2 version 6, markdown formating can be used in the documentation, 
by specifying the `@md` tag.

## Unit tests
Back in R, add testing infrastructure:

    devtools::use_testthat()

When checking the package with R CMD CHECK, 
[How can I handle R CMD check “no visible binding for global variable”?](https://stackoverflow.com/questions/9439256/how-can-i-handle-r-cmd-check-no-visible-binding-for-global-variable-notes-when) 
These notes are caused by variables used with dplyr verbs and ggplot2 aesthetics.


## Continuous Integration
It is good to know if your package can be installed on a fresh system. 
Continuous integration systems make this possible each time you submit a modification to your repository. 
I have used travis-ci which is free for open github repositories. 
[Instructions to build an R project on travis](https://docs.travis-ci.com/user/languages/r/).
Unit tests are also run on travis, in addition to R CMD CHECK.

Package dependencies can be configured in a `.travis.yml`
file that is read by the travis machine performing the build.
For package that are not on Cran, it's possible to specify a dependency field under `r_github_packages`.

