---
title: "R Commands"
author: "Paul Rougieux"
date: "17 August 2017"
output: 
  html_document: 
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


This page is the continuation of [my blog post on R commands](https://paulremote.blogspot.de/2014/04/r-commands.html). 
On the blog, see also [why use R](http://paulremote.blogspot.fr/2013/10/why-r.html) and the [RSS feed of posts labelled R](http://paulremote.blogspot.com/feeds/posts/default/-/R).

See also documentation at:

 * The Comprehensive R Archive Network [manuals](https://cran.r-project.org/manuals.html)
 * Tidyverse [packages](http://www.tidyverse.org/packages/)
 


# Set operations

```{r}
x = letters[1:3]
y = letters[3:5]
union(x, y)
intersect(x, y)
setdiff(x, y)
setdiff(y, x)
setequal(x, y)
```



# Information about your R system
```{r eval=FALSE}
sessionInfo()
installed.packages()
```


# Handling files

    getwd()
    list.files(tempdir()) 
    dir.create("blabla")
    read.csv("data.csv")

# Lists
Given a list structure x, unlist simplifies it to produce a vector which contains all the atomic components which occur in x.
```{r}
l1 <- list(a="a", b="2,", c="pi+2i")
str(l1)
unlist(l1) # a character vector 
str(unlist(l1))
```


# Strings

```{r}
message("Using the following letters: ", paste(letters, collapse=","), ".")
```

# S3 methods
x<-1



List all available methods for a class:

    methods(class="lm") 


# One liners
Remove all objects in the workspace except one :

    rm(list=ls()[!ls()=="object_to_keep"]) 


# knitr
Those 2 commands are different.
Sets the options for chunk, within a knitr chunk inside the .Rmd document

    opts_chunk$set(fig.width=10)

 Sets the options for knitr outside the .Rmd document

    opts_knit$set()


# dplyr
pipes

```{r warning=FALSE, message=FALSE}
library(dplyr)
cars %>%
    group_by(speed) %>%
    print() %>% # works because the print function returns its argument
    summarise(numberofcars = n(),
              min = min(dist),
              mean = mean(dist),
              max = max(dist)) 
```



group_by() creates a tbl_df objects which is a wrapper around a data.frame to enable some functionalities. Note that print returns its output on a tbl_df object. So print() can be used inside the pipe without stopping the workflow.


# tidyr

[tidyr vignette on tidy data](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html) 
In the section on "Multiple types in one table":

> Datasets often involve values collected at multiple levels, on different types of observational units. During tidying, each type of observational unit should be stored in its own table. This is closely related to the idea of database normalisation, where each fact is expressed in only one place. It’s important because otherwise inconsistencies can arise.

> Normalisation is useful for tidying and eliminating inconsistencies. However, there are few data analysis tools that work directly with relational data, so analysis usually also requires denormalisation or the merging the datasets back into one table.



# Plotting with ggplot2

    geom_bar
    geom_tile + a gradient produce heat maps

