---
title: "R Commands"
author: "Paul Rougieux"
date: "17 August 2017"
output: 
  html_document: 
    toc: yes
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This page is the continuation of [my blog post on R commands](https://paulremote.blogspot.de/2014/04/r-commands.html). 
On the blog, see also [why use R](http://paulremote.blogspot.fr/2013/10/why-r.html) and the [RSS feed of posts labelled R](http://paulremote.blogspot.com/feeds/posts/default/-/R).

See also documentation at:

 * The Comprehensive R Archive Network CRAN [manuals](https://cran.r-project.org/manuals.html)
 * The Tidyverse [packages](http://www.tidyverse.org/packages/)


# Control flow

dplyr::case_when is a generalised vectorised if. Example use:

     x <- 1:50
     case_when(
       x %% 35 == 0 ~ "fizz buzz",
       x %% 5 == 0 ~ "fizz",
       x %% 7 == 0 ~ "buzz",
       TRUE ~ as.character(x)
     )


# Data table

[Database like operation benchmark](https://h2oai.github.io/db-benchmark/#task)
between data table, dplyr, python pandas and other database like tools.


# Data types

## Everything is a vector

There are no scalar values in R, only vectors, this makes it very natural to
process times series or cross sectional data over many N observations as if we
were processing only one instance.

[SO answer concerning indexing up to end of vector/matrix](https://stackoverflow.com/a/7500961/2641825) "Sometimes it's easier to tell R what you don't want"

    x <- c(5,5,4,3,2,1)
    x[-(1:3)]
    x[-c(1,3,6)]

### Set operations

```{r}
x = letters[1:3]
y = letters[3:5]
union(x, y)
intersect(x, y)
setdiff(x, y)
setdiff(y, x)
setequal(x, y)
```

## Factors

[My answer to the question how to reorder factor levels the tidy
way](https://stackoverflow.com/questions/45148897/how-to-reorder-factor-levels-in-a-tidy-way/59246372#59246372) 

If you happen to have a character vector to order, for example:

    iris2 <- iris %>%
        mutate(Species = as.character(Species)) %>%
        group_by(Species) %>%
        mutate(mean_sepal_width = mean(Sepal.Width)) %>%
        ungroup()

You can also order the factor level using the behavior of the forcats::as_factor function :

    "Compared to base R, this function creates levels in the order in which they appear"

    library(forcats)
    iris2 %>%
        # Change the order
        arrange(mean_sepal_width) %>%
        # Create factor levels in the order in which they appear
        mutate(Species = forcats::as_factor(Species)) %>%
        ggplot() +
        aes(Species, Sepal.Width, color = Species) +
        geom_point()

Notice how the species names on the x axis are not ordered alphabetically but
by increasing value of their mean_sepal_width. Remove the line containing
as_factor to see the difference. 

## Lists

Given a list structure x, unlist simplifies it to produce a vector which contains all the atomic components which occur in x.
```{r}
l1 <- list(a="a", b="2,", c="pi+2i")
str(l1)
unlist(l1) # a character vector 
str(unlist(l1))
```

Extract the first element of a nested list [SO](https://stackoverflow.com/questions/20428742/select-first-element-of-nested-list)

```{r}
minyearmaxyear <- list(c(2001, 2009), c(2010, 2014), c(2015, 2100))
library(purrr)
map(minyearmaxyear, 1)
```


## Strings

```{r}
message("Using the following letters: ", paste(letters, collapse=","), ".")
```

### Levenshtein distance between words

Cf. https://en.wikipedia.org/wiki/Levenshtein_distance

    adist("kitten", "sitting")

## S3 methods

List all available methods for a class:

    methods(class="lm") 


# Debug

## Debug with base R

Print the call stack of the last error:

    traceback()


## Debug with rlang

Print the call stack of the last error in simplified form:

    rlang::last_error()

Print the full call stack of the last error:

    rlang::last_trace()

More information on the behaviour of backtrace:

    help(trace_back, package=rlang)

> "Because of lazy evaluation, the call stack in R is actually a tree, which
> the ‘summary()’ method of this object will reveal." 


# Environement

List objects in the global environment:

    ls

List objects coming from a specific loaded package:

    ls("package:stats")
    ls("package:stats", pattern="smooth")
 
Remove all objects in the environement except one:

    rm(list=ls()[!ls()=="object_to_keep"]) 
    rm(list=ls()[!ls()=="con"]) # Remove all except a database connection

Get the current environment

    rlang::current_env()


# Editors

I use Vim and the great [Nvim-R](https://github.com/jalvesaq/Nvim-R) plugin to
edit R markdown documents and R scripts. I browse the table of content of R
markdown files with the Voom plugin. I create R packages with Makefiles. I use
the fugitive plugin to manage my git commits. See my [Vim page](Vim.html) for more
information. 

Rstudio is a great editors for those who prefer a graphical user interface.
It has a Vim mode. In fact I trained my muscle memory for Vim movement keys inside
RStudio initially. Rstudio also has a lot of menus and buttons to work with
Rmarkdown documents, view the state of a git repository and build packages.


# Input output

    getwd()
    list.files(tempdir()) 
    dir.create("blabla")

## CSV files

Read one csv file with default R function.

    read.csv("data.csv", )

Read many csv files with functions from the tidyverse packages.

First write sample csv files to a temporary directory.
It's more complicated than I thought it would be.

    data_folder <- file.path(tempdir(), "iris")
    dir.create(data_folder)
    iris %>%
        # Keep the Species column in the output
        # Create a new column that will be used as the grouping variable
        mutate(species_group = Species) %>% 
        group_by(species_group) %>%
        nest() %>%
        by_row(~write.csv(.$data, 
                          file = file.path(data_folder, paste0(.$species_group, ".csv")),
                          row.names = FALSE))


Read these csv files into one data frame.
Note the Species column has to be present in the csv files, otherwise we would loose that information. 

    iris_csv <- list.files(data_folder, full.names = TRUE) %>% 
        map_dfr(read_csv)

`write_csv` returned an `Error in write_delim(...) is.data.frame(x) is not TRUE`
That's why we used `write.csv` instead.


## RDS files

RDS is a Serialization Interface for Single Objects. Objects are saved to
binary data and compressed to the gzip format, see `help(saveRDS)` for more
details.

Write a dataset to an rds file:

    saveRDS(iris, "/tmp/iris.rds")

Read a dataset from an rds file

    iris2 <- readRDS("/tmp/iris.rds")
    identical(iris, iris2)

## Databases

Dplyr has databases backends  to MySQl (MariaDB) and PostgreSQL.



# Find and replace

## Longest contiguous stretch of non-NAs:

    na.contiguous(c(NA,1:3,NA,NA,3:5,NA,NA))
    ] 1 2 3
    tr(,"na.action")
    ]  1  5  6  7  8  9 10 11
    tr(,"class")
    ] "omit"
    tr(,"tsp")
    ] 2 4 1
    na.contiguous(c(NA,1:3,NA,NA,3:6,NA,NA))
    ] 3 4 5 6
    tr(,"na.action")
    ]  1  2  3  4  5  6 11 12
    tr(,"class")
    ] "omit"
    tr(,"tsp")
    ]  7 10  1

## Replace text in column names

Add the first row to the column name, useful for some double headed csv files
coming from pandas.

    paste(names(df), df[1,])

Replace a regular expression with optional numerical pattern and final space

    names(df) <- gsub(".per.hectare.[0-9]* *", "_", names(df))

[And in a regular
expression](https://stackoverflow.com/questions/18237852/grep-in-r-using-or-and-not/18239222)

    |


# Installing R and packages

To install R and Rstudio on Debian, see the 
[debian.html#r_and_rstudio](debian.html#r_and_rstudio) page on this site.

To install packages simply enter the following at an R command prompt:

    install.packages("package_name")

Some packages have dependencies that need to be installed at the OS level.
Error messages:

    "Configuration failed because libcurl was not found." 
    "Configuration failed because libxml-2.0 was not found."
    "Configuration failed because openssl was not found."

Can be solved by installing these dependencies: 

    sudo apt install libcurl4-openssl-dev 
    sudo apt install libxml2-dev 
    sudo apt install libssl-dev

# Information about your R system

```{r eval=FALSE}
sessionInfo()
installed.packages()
```

# knitr


## Create websites

### Rmarkdown site generator

This website is generated from R Markdown documents with: 

	 Rscript -e "rmarkdown::render_site()"

The advantage of the rmarkdown site generator is that it creates pages with a
floating, foldable table of content. It's nice for pages that have a lot of
subdivisions like this blog. The structure of a website is specified in a
yaml file. For example, the code below creates a navigation bar:

    navbar:
      title: "Paul Rougieux"
      left:
        - text: "Home"
          icon: fa-home 
          href: index.html
        - text: "Tools"
          icon: fa-wrench
          menu:
            - text: "Bash"
              href: bash.html


### Bookdown site generator

For blogs that fit into a book structure, with chapters, the bookdown package
create a single clickable table of content for the whole site. It can also
export the content to pdf format.

Specifying input from many sub directories in `_bookdown.yml`:

    rmd_files: ["index.Rmd", "chapters/chapt1.RMD", "chapters/chapt2.RMD"]

Attention this will only work in the Merge and Knit strategy, as explained in
the [book chapter on two rendering
approaches](https://bookdown.org/yihui/bookdown/new-session.html) 

> "We call these two approaches “Merge and Knit” (M-K) and “Knit and Merge”
> (K-M), respectively." "K-M does not allow Rmd files to be in subdirectories,
> but M-K does."


## Cross references

[R markdown cross
ref](https://bookdown.org/yihui/rmarkdown-cookbook/cross-ref.html)

> "Cross-referencing is not provided directly within the base rmarkdown
> package, but is provided as an extension in bookdown (Xie 2020c). We must
> therefore use an output format from bookdown (e.g., html_document2,
> pdf_document2, and word_document2, etc.) in the YAML output field."

* Bookdown issue with flextable [Table and figure cross references in word](https://github.com/rstudio/bookdown/issues/746)
Now solved with the following implementation

```{r irishead, ft.align="left", tab.cap="Edgar Anderson's Iris Data", eval=FALSE}
iris %>% head() %>% flextable()
```
This is a reference to flextable \@ref(tab:irishead).
The chunk name should not contain an underscore, otherwise the cross reference
will not work. See [bookdown issue 941](https://github.com/rstudio/bookdown/issues/941).

See more flextable caption examples in `system.file(package = "flextable", "examples", "rmd", "captions")`


## Flextable and officer package

The officer package makes it possible to create improved word and powerpoint
documents. The flextable package has a few functions to generate tables for
word documents (otherwise they look funny).

Before installing flextable, the following system dependencies were required:

    sudo apt install libcairo2-dev libjpeg-dev libgif-dev

Installation in R:

    install.packages("flextable")

This is to work with colleagues who are not in the latex/pdf world.

### Landscape pages
```{r insert_landscape_section, eval=FALSE}
library(officer)
library(flextable)
ft_iris <- iris %>%
    head() %>%
    flextable() %>%
    autofit()
# Add a flex table in landscape format 
# Supply document name here, otherwise start an empty doc
doc <- read_docx() %>% 
  # Portait and landscape sections are defined by their ending
  # A portrait section is ending here
  body_end_section_portrait() %>%
  body_add_flextable(value = ft_iris, split = TRUE) %>%
  # A landscape section is ending here
  body_end_section_landscape() %>% 
  print(target = "/tmp/iris.docx")
```


## kable to create tables

```{r }
cat(kable(head(iris, 1), format = "html"))
cat(kable(head(iris, 1), format = "latex"))
cat(kable(head(iris, 1), format = "markdown"))
```


### Booktabs

Booktabs tables look nicer in latex document, they can be created by specifying
the argument `booktabs = TRUE` into the `kable()` call.

Knitr adds an extra space every five lines into booktabs tables for
readability. This is undesirable for short tables. To remove this space specify
the `linesep = ""` argument to `kable()` as explained in this [SO
answer](https://stackoverflow.com/questions/49015578/space-after-every-five-rows-in-kable-output-with-booktabs-option-in-r-markdown). 


## Python engine

Rstudio: [R Mardown python engine](https://rstudio.github.io/reticulate/articles/r_markdown.html)


## Set knitr options
Those 2 commands are different.
Sets the options for chunk, within a knitr chunk inside the .Rmd document

    opts_chunk$set(fig.width=10)

 Sets the options for knitr outside the .Rmd document

    opts_knit$set()

# Manipulate vectors

## Cut and split

Cut a vector in smaller components

    cut(0:16, 4)


## Combine vectors

Combine vectors with c()

    x <- c(1, 2, 3)
    y <- c(4, 5)
    c(x, y)

Unite multiple columns into one by pasting strings together

    df <- expand_grid(x = c("a", NA), y = c("b", NA))
    df
    
    df %>% unite("z", x:y, remove = FALSE)
    # To remove missing values:
    df %>% unite("z", x:y, na.rm = TRUE, remove = FALSE)
    
    # Separate is almost the complement of unite
    df %>%
      unite("xy", x:y) %>%
      separate(xy, c("x", "y"))
    # (but note `x` and `y` contain now "NA" not NA)


# Packages
You might want to read the CRAN manual on [Writing R Extensions](https://cran.r-project.org/doc/manuals/R-exts.html), and its section on [Package dependencies](https://cran.r-project.org/doc/manuals/R-exts.html#Package-Dependencies).
See also Hadley's book on R package and its section on [Namespace](http://r-pkgs.had.co.nz/namespace.html)

Use the devtools library to start a package folder structure:

    devtools::create("package_name")

Use git to track code modifications (shell commands):

    $ cd package_name
    $ git init

## Documentation

The roxygen2 package helps with function documentation.
Documentation can be written in the form of comments #'
tags such as `@param` and `@description` structure the documentation of each function.

For an introcution to roxygen2, call
`vignette("roxygen2", package = "roxygen2")` at the R prompt.

Since roxygen2 version 6, markdown formating can be used in the documentation, 
by specifying the `@md` tag.

### Examples
Examples are crucial to demonstrate the use of a fonction. 
They can be specified in a roxygen block called `@examples`:

	#' @examples

Wrap the examples in donttest if you don't want R CMD check to test them at package building time.

	#' \donttest{d

It is also possible to wrap them in another statement called `dontrun`, but this is not recomended on CRAN
according to this [Stackoverflow question](https://stackoverflow.com/a/44997778/2641825).

### Vignettes

[Vignettes: long-form documentation](http://r-pkgs.had.co.nz/vignettes.html)

    devtools::use_vignette("my-vignette")

[Where to put package vignettes for CRAN submission](https://stackoverflow.com/questions/12325223/where-to-put-package-vignettes-for-cran-submission)

> "You put the .Rnw sources in vignettes/ as you did, but you missed out a critical step; don't check the source tree. The expected workflow is to build the source tarball and then check that tarball. Building the tarball will create the vignette PDF."

    R CMD build ../foo/pkg
    R CMD check ./pkg-0.4.tar.gz

Issues while building vignette for a packages:

* sh: 1: /usr/bin/texi2dvi: not found

    sudo apt-get install texinfo

* [Warning about "vignette without corresponding PDF/HTML"?](https://github.com/clonghurst/Res.plotter/issues/3)

> "Maybe you're running R CMD check using the directory name rather than the .tar.gz file?"

* [Inconsolata missing to build R vignette](https://stackoverflow.com/questions/34524357/inconsolata-missing-to-build-r-vignette)

> "Installing texlive-fonts-extra should take care of it."


R CMD checking data for non-ASCII characters found 179 marked UTF-8 strings 
No solution for this one but I guess it's ok since it concerns country names? 

## Unit tests
Back in R, add testing infrastructure:

    devtools::use_testthat()

When checking the package with R CMD CHECK, 
[How can I handle R CMD check “no visible binding for global variable”?](https://stackoverflow.com/questions/9439256/how-can-i-handle-r-cmd-check-no-visible-binding-for-global-variable-notes-when) 
These notes are caused by variables used with dplyr verbs and ggplot2 aesthetics.


## Continuous Integration
It is good to know if your package can be installed on a fresh system. 
Continuous integration systems make this possible each time you submit a modification to your repository. 
I have used travis-ci which is free for open github repositories. 
[Instructions to build an R project on travis](https://docs.travis-ci.com/user/languages/r/).
Unit tests are also run on travis, in addition to R CMD CHECK.

Package dependencies can be configured in a `.travis.yml`
file that is read by the travis machine performing the build.
For package that are not on Cran, it's possible to specify a dependency field under `r_github_packages`.

## Make file

* kbroman [minimal make](https://kbroman.org/minimal_make/)
* Cran [how to use make files with R CMD build](https://stackoverflow.com/questions/46741739/how-to-use-makefiles-with-r-cmd-build)

I copied the make file from the knitr package.

## Submitting to CRAN

[Submit package to CRAN](https://cran.r-project.org/submit.html)
The second page will ask to 

* Step 1 (Upload) the `*.tar.gz` file.
* Step 2 (Submission) review information from the package's DESCRIPTION file. 
* Step 3 (Confirmation)

The maintainer of this package has been sent an email to confirm the submission. After their confirmation the package will be passed to CRAN for review. 

In general enter the Git tag after the submission has been completed, because
it is very likely that additional information will need to be added to the
package before or during the submission process.

# Plots


## Plotting with ggplot2

    geom_bar
    geom_tile + a gradient produce heat maps

### Axes

Rotate the tick text of the X axis. SO question [rotating axis labels in
ggplot2](https://stackoverflow.com/questions/1330989/rotating-and-spacing-axis-labels-in-ggplot2) 

    theme(axis.text.x=element_text(angle = 90, hjust=0))

Increase number of axis ticks by specifying an integer break vector

    scale_x_continuous(breaks = 1:10)

When dates are on the x axis, use a time scale and specify the breaks as a time vector

    scale_x_date(breaks = lubridate::parse_date_time(2000:2020, "%y"))

Alternatively breaks can be specified as an interval, and minor breaks as well

    scale_x_datetime(date_breaks = "5 years", labels = scales::date_format("%Y"), date_minor_breaks = "1 year")

### Legend

Place the legend at the bottom of the plot:

    library(ggplot2)
    ggplot(iris, aes(Petal.Width, Sepal.Width, color=Species)) + 
        geom_point() + 
        theme(legend.position="bottom")

Change the name of a particular sub element in the legend:

    labs(color="Iris Species")
    labs(linetype="Sawnwood product group")

Make a multi column legend

    library(scales)
    
    plot + # With a fill aeasthetic
        guides(fill=guide_legend(ncol = 2))

    plot + # with a color aesthetic 
         guides(col = guide_legend(ncol = 1))

See ggplot2 reference [guide
legend](https://ggplot2.tidyverse.org/reference/guide_legend.html).


### Math formulas in facet labels

```{r}
library(dplyr)
library(ggplot2)

# Create a first facet variable with examples of math formulas
iris2 <- iris %>%
    mutate(species_math = factor(Species,
                            levels = c("setosa", "versicolor", "virginica"),
                            labels = c("m^2",
                                       expression(bar(x) == sum(frac(x[i], n), i==1, n) * beta * Q[t-1]),
                                       bquote(pi == .(pi)))))

# Create a second facet variable with mean lengths
# This illustrates how to pass a numeric vector inside a formula
iris_mean <- iris2 %>%
    group_by(Species) %>%
    summarise(across(ends_with("Length"), mean), .groups="drop")

iris2$mean_length <- factor(iris2$Species,
                           levels =  c("setosa", "versicolor", "virginica"),
                           labels = mapply(function(p, s) bquote(bar(p) == .(p) ~ bar(s) ==.(s)),
                                            round(iris_mean$Petal.Length,3), round(iris_mean$Sepal.Length,3)))

# Create the faceted plot
iris2 %>%
    ggplot(aes(x = Petal.Length, y = Petal.Width)) +
    geom_point() +
    facet_wrap(species_math ~ mean_length + Species,
               labeller = labeller(species_math = label_parsed, mean_length = label_parsed))

```

As shown in the example above, the labeller can parse:

1. A character vector such as "m^2" for simple formulas

2. An `expression` for more complex math with indices

3. The output of `bquote` to include numerical values in the formula. See also
   [this answer](https://stackoverflow.com/a/58048944/2641825) on how to use
   `bquote` with numerical vectors of more than one value. 

4. see [this other answer](https://stackoverflow.com/a/51863090/2641825) on how
   to  apply the labeller to one of the faceting variables only. In our case we
   apply it to the `species_math` variable only. 

The syntax is different from Latex math formulas because `label_parsed`
interprets labels as `plotmath` expressions. For example indices are written
`x_i` in Latex and `x[i]` in plot math expressions, and Greek letters are
written directly as `alpha` instead of `\alpha` in Latex. You can find many
formulas in the help page of the
[plotmath](https://stat.ethz.ch/R-manual/R-devel/library/grDevices/html/plotmath.html)
function. Good luck with the `plotmath` examples. 


### Save plots to pdf


Use [ggsave](https://ggplot2.tidyverse.org/reference/ggsave.html):

    ggplot(mtcars, aes(mpg, wt)) + geom_point()
    ggsave("mtcars.pdf")

### Themes

Minimal theme

    theme_minimal()


## Choropleth maps

[Choropleth map](https://www.r-graph-gallery.com/choropleth-map.html)

Shape files to make choropleth maps of Europe:
[GISCO: Geographical Information and maps](https://ec.europa.eu/eurostat/web/gisco/geodata/reference-data/administrative-units-statistical-units/nuts)



## Palettes

[Setting up colour palettes in R](https://data.library.virginia.edu/setting-up-color-palettes-in-r/)

> To create a RColorBrewer palette, use the brewer.pal function. It takes two
> arguments: n, the number of colors in the palette; and name, the name of the
> palette. Let’s make a palette of 8 colors from the qualitative palette,
> “Set2”. 


    library(RColorBrewer)
    brewer.pal(n = 8, name = "Set2") 
    [1] "#66C2A5" "#FC8D62" "#8DA0CB" "#E78AC3"
    "#A6D854" "#FFD92F" "#E5C494" "#B3B3B3" 
    palette(brewer.pal(n = 8, name = "Set2"))

Use this palette in ggplot2

    ggplot(iris, aes(x=Sepal.Length, y=Petal.Length, color=Species)) + 
        geom_point() +
        scale_color_brewer(palette = "Set2")

### Manual colours

Use a named vector to set a palette in ggplot2 as explained in 
[ggplot2 scale_manual](https://ggplot2.tidyverse.org/reference/scale_manual.html)

    p <- ggplot(iris, aes(x=Sepal.Length, y=Petal.Length, color=Species)) +
            geom_point() 
    p + scale_colour_manual(values = c(setosa='black', versicolor='red', virginica='green'))

Create a named palette using R colour brewer:

    species_names <- c("setosa", "versicolor", "virginica") 
    iris_palette <- setNames(brewer.pal(n=length(species_names), name='Set2'), 
                             species_names)
    p + scale_colour_manual(values = iris_palette)

### Display palettes

Display qualitative palettes:

    display.brewer.all(type="qual") 

Display all palettes 

    display.brewer.all()

## Text labels

ggplot2 [text](https://ggplot2.tidyverse.org/reference/geom_text.html)

Use position jitter

```{r eval=FALSE}
#  Reduce the number of points so it becomes more readable
iris2 <- iris %>% 
    group_by(Species, Petal.Width) %>%
    summarise(Sepal.Width = mean(Sepal.Width))
ggplot(iris2, aes(x = Petal.Width, y=Sepal.Width, color = Species)) +
    geom_text(aes(label = Species), position=position_jitter(width=0.5,height=0.3))
```

Use repulsive textual annotations

```{r eval=FALSE}
ggplot(iris2, aes(x = Petal.Width, y=Sepal.Width, color = Species)) +
    ggrepel::geom_text_repel(aes(label = Species))
```

Change the text size

```{r eval=FALSE}
ggplot(iris2, aes(x = Petal.Width, y=Sepal.Width, color = Species)) +
    ggrepel::geom_text_repel(aes(label = Species), size=8)
```

[More ggrepel examples](https://ggrepel.slowkow.com/articles/examples.html)

Justification

```{r eval=FALSE}
df <- data.frame(
  x = c(1, 1, 2, 2, 1.5),
  y = c(1, 2, 1, 2, 1.5),
  text = c("bottom-left", "bottom-right", "top-left", "top-right", "center")
)
ggplot(df, aes(x, y)) +
  geom_text(aes(label = text))
ggplot(df, aes(x, y)) +
  geom_text(aes(label = text), vjust = "inward", hjust = "inward")
```


# Python compared to R

See also the [python](python.html) page for a comparison between pandas methods
and R data frames functions.

    l = c(1,2,3)
    s = l
    s[3]
    [1] 3
    s[3] = "a"
    s
    [1] "1" "2" "a"
    l
    [1] 1 2 3

Using the address function to see the address of these objects in memory
We can see that s and l share the same address. 
The address only changes when we asign something to s. 

    library(pryr)
    l = c(1,2,3)
    s = l
    address(l)
    [1] "0x316f718"
    address(s)
    [1] "0x316f718"
    s[3] = "a"
    address(s)
    [1] "0x36a7d30"
    s
    [1] "1" "2" "a"
    l
    [1] 1 2 3

Checking string objects

    bla = "qsdfmlkj"
    address(bla)
    [1] "0x38e5120"
    bli = bla
    address(bli)
    [1] "0x38e5120"
    bli = paste(bli, "sdf")
    address(bli)
    [1] "0x38d1d60"

TODO Compare to the same code in python
to see the difference between the above and passing by reference.


# Reproducibility and cloud computing

## Cloudyr project

The [cloudyr project](https://cloudyr.github.io/packages/index.html) is a
collection of R packages to enable cloud computing. 

R packages that are actively maintained can be seen on the [github project page of
cloudyr](https://github.com/cloudyr).

- For example the
  [googleCloudVisionR](https://github.com/cloudyr/googleCloudVisionR) package
  gets the following image annotations for a picture of golden retriever
  puppies

          description     score topicality
                  Dog 0.9953705  0.9953705
               Mammal 0.9890478  0.9890478
           Vertebrate 0.9851104  0.9851104
              Canidae 0.9813780  0.9813780
            Dog breed 0.9683250  0.9683250
                Puppy 0.9400384  0.9400384
     Golden retriever 0.8966703  0.8966703


## Rocker

[rocker-org/rocker](https://github.com/rocker-org/rocker) provides a series of
docker images for various R development purposes.

Bioconductor [docker intro](https://www.bioconductor.org/help/docker/#intro)

> "With Bioconductor containers, we hope to enhance Reproducibility: If you run
> some code in a container today, you can run it again in the same container
> (with the same tag) years later and know that nothing in the container has
> changed. You should always take note of the tag you used if you think you
> might want to reproduce some work later."

# Rscript

Capture arguments in an Rscript on windows and write them to a file
```
"C:\Program Files\R\R-3.5.0\bin\Rscript.exe" --verbose -e "args = commandArgs(trailingOnly=TRUE)" -e "writeLines(args,'C:\\Dev\\args.txt')" "file1.csv" "file2.csv" "file3.csv"
```
Arguments can be extracted one by one with args[1]
commandArgs() returns a **character vector** containing the name of the executable and the user-supplied command line arguments.


# Shiny

[K means example](https://shiny.rstudio.com/gallery/kmeans-example.html)

## App.R

[Shiny tutorial lesson 1](https://shiny.rstudio.com/tutorial/written-tutorial/lesson1/)

> "Note: Prior to version 0.10.2, Shiny did not support single-file apps and
> the ui object and server function needed to be contained in separate scripts
> called ui.R and server.R, respectively. This functionality is still supported
> in Shiny, however the tutorial and much of the supporting documentation focus
> on single-file apps."



# Tidyverse


## dplyr

### Group_by and summarise

```{r warning=FALSE, message=FALSE}
library(dplyr)
cars %>%
    group_by(speed) %>%
    print(n=2) %>% # works because the print function returns its argument
    summarise(numberofcars = n(),
              min = min(dist),
              mean = mean(dist),
              max = max(dist)) 
```

group_by() creates a tbl_df objects which is a wrapper around a data.frame to enable some functionalities. Note that print returns its output on a tbl_df object. So print() can be used inside the pipe without stopping the workflow.


### Across

Apply the same transformation to multiple columns, allowing to use select
semantics inside summarise and mutate.

```{r}
cars %>%
    group_by(speed) %>%
    summarise(across(dist, list(n = length , min = min, mean = mean, max = max)))
```

Other examples from the help of the across function

```{r}
# A purrr-style formula
iris %>%
    group_by(Species) %>%
    summarise(across(starts_with("Sepal"), ~mean(.x, na.rm = TRUE)))

# A named list of functions
iris %>%
    group_by(Species) %>%
    summarise(across(starts_with("Sepal"), list(mean = mean, sd = sd)))
```

More information in the vignette

    vignette("colwise", package="dplyr")

The vignette mentions this about `rename`: 

> "across() doesn’t work with select() or rename() because they already use
> tidy select syntax; if you want to transform column names with a function,
> you can use rename_with()." 


### Mutate to create a new vector

Mutate multiple variables in the dataframe at once using the `vars()` helper function to scope the mutation:
Note `mutate_at` has been superseded by the across function.

```{r} 
iris %>%
    mutate_at(vars(starts_with("Petal")), round) %>%
    head()
```


### Rename columns

`rename_with` from the `dplyr` package can use either a function or a formula
to rename a selection of columns given as the `.cols` argument. For example
passing the function name `toupper`: 

    library(dplyr)
    rename_with(head(iris), toupper, starts_with("Petal"))

Is equivalent to passing the formula `~ toupper(.x)`:

    rename_with(head(iris), ~ toupper(.x), starts_with("Petal"))

To rename all columns, you can also use `set_names` from the rlang package. We
now use `paste0` as a renaming function. `pasteO` takes 2 arguments, as a
result there are different ways to pass the second argument depending on
whether we use a function or a formula. 

    rlang::set_names(head(iris), paste0, "_hi")
    rlang::set_names(head(iris), ~ paste0(.x, "_hi"))

The same can be achieved with `rename_with` by passing the data frame as first
argument `.data`, the function as second argument `.fn`, all columns as third
argument `.cols=everything()` and the function parameters as the fourth
argument `...`. Alternatively you can place the second, third and fourth
arguments in a formula given as the second argument.

    rename_with(head(iris), paste0, everything(), "_hi")
    rename_with(head(iris), ~ paste0(.x, "_hi"))

`rename_with` only works with data frames. `set_names` is more generic and can
also perform vector renaming

    rlang::set_names(1:4, c("a", "b", "c", "d"))

### Recode values in a vector

> "Replace numeric values based on their position or their name, and character
> or factor values only by their name." 

    char_vec <- sample(c("a", "b", "c"), 10, replace = TRUE)
    recode(char_vec, a = "Apple")


## Pipes


## purrr

Hadley Wickham's answer to a SO question 
[Why use purrr::map instead of lapply?](https://stackoverflow.com/a/47123420/2641825)


### Map a function to nested data sets

Load data

```{r eval=FALSE}
list.files(getwd())
forestEU_wide <- read.csv("Forest-R-EU.csv", stringsAsFactors = FALSE)
head(forestEU)
```
Pivot to long format
```{r eval=FALSE}
# at some pointin the future this will be called pivot_long
forestEU <- forestEU_wide  %>% 
      # select everything except the Year, then pivot all columns and put the value in area
      gather(-Year, key = "Country", value = "Area")
```

## Tricks

David Robinson [Ten tricks in the tidyverse](https://www.youtube.com/watch?v=NDHSBUN_rVU)

1. count with weight, sort and renamining 
   count(x, ..., wt = NULL, sort = FALSE, name = "n")
2. creating variables in count()
3. add_count()
4. summarize with a list column (to create many models for example)
5. fct_reorder() + geom_col() + coord_flip()
6. fct_lump() to lump less frequent values together
7. use a log scale scale_x/y_log10
8. crossing()
9. separate()
10. extract() use a regular expression such as `S(.*)E(.*)` to extract series
    and episode from a string of the form "S32E44".


## Purrr

### Interpolation

Interpolate for one country

```{r eval=FALSE}
country_interpolation <- function(df) {
    df <- data.frame(approx(df$Year, df$Area, method = "linear", n=71))
    df <- rename(df, Year = x, Area = y)
    return(df)
}
forestEU %>%
    filter(Country=="Austria") %>%
    country_interpolation()
```


Interpolate for all countries
Perform the Interpolation on all countries


See documentation in 
* many models https://r4ds.had.co.nz/many-models.html
* blog https://emoriebeck.github.io/R-tutorials/purrr/


```{r eval=FALSE}
forestEU_nested <- forestEU %>%
      # Remove empty area
      filter(!is.na(Area)) %>% 
      group_by(Country) %>%  
      nest() %>% 
      mutate(interpolated = map(data, country_interpolation))

# forestEU_nested %>% unnest(data)
# Unnest the interpolated data to look at it and plotting
forestEU_interpolated <- forestEU_nested %>% unnest(interpolated)
```


### Read many csv

See iris_csv file created in the csv section of this document. Read these csv
files into one data frame. Note the Species column has to be present in the csv
files, otherwise we would loose that information. 

```{r eval=FALSE}
iris_csv <- list.files(data_folder, full.names = TRUE) %>% 
    purrr::map_dfr(read_csv)
```


### Write to many csv

```{r eval=FALSE}
#dir.create("output")
forestEU_nested <- forestEU_nested %>% 
    mutate(filename = paste0("output/", Country, ".csv"),
    wrote_stuff = map2(interpolated, filename, write.csv))
```


## tidy evaluation

* Advanced R [metaprogramming section](https://adv-r.hadley.nz/meta.html)
* [Examples of tidy evaluation with plots](https://www.tidyverse.org/articles/2018/07/ggplot2-tidy-evaluation/)

```{r eval = FALSE}
scatter_plot <- function(data, x, y) {
    x <- enquo(x)
    y <- enquo(y)
    ggplot(data) + geom_point(aes(!!x, !!y))
}
scatter_plot(mtcars, disp, drat)
```

Another example use of metaprogramming to change variables
```{r eval = FALSE}
add1000 <- function(dtf, var){
      varright <- enquo(var)
  varleft <- quo_name(enquo(var))
    dtf %>% 
            mutate(!!varleft := 1000 + (!!varright))
}
add1000(iris, Sepal.Length)
```


* [Tidy evaluation explained in French](https://thinkr.fr/tidyeval/)


## Tidy select

Tidy select [selection
language](https://tidyselect.r-lib.org/reference/language.html) explains the
various selection function such as `all_of`, `contains` and `starts_with`.

```{r eval=FALSE}
library(dplyr)
iris2 <- head(iris, 1)
select(iris2, contains("w")) # Literal string
select(iris2, matches("w")) # Regular expression
select(iris2, starts_with("S"))
select(iris2, all_of(c("Sepal.Length", "Petal.Width")))
```
The selection can be inverted by using s minus sign in front of it. Not it's
not a question mark as explained in this  [SO
answer](https://stackoverflow.com/a/45941792/2641825) because the selection
fonctions return numerical vectors of integer column positions (not boolean
values).

```{r eval=FALSE}
library(dplyr)
select(head(iris), -matches("w") )
```


## tidyr

[tidyr vignette on tidy data](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html) 
In the section on "Multiple types in one table":

> Datasets often involve values collected at multiple levels, on different types of observational units. During tidying, each type of observational unit should be stored in its own table. This is closely related to the idea of database normalisation, where each fact is expressed in only one place. It’s important because otherwise inconsistencies can arise.

> Normalisation is useful for tidying and eliminating inconsistencies. However, there are few data analysis tools that work directly with relational data, so analysis usually also requires denormalisation or the merging the datasets back into one table.

Example use of `tidyr::nest()` to generate a group of plots: 
[make ggplot2 purrr](http://www.brodrigues.co/blog/2017-03-29-make-ggplot2-purrr/).

```{r eval=FALSE}
library(tidyr)
library(dplyr)
library(purrr)
library(ggplot2)
piris <- iris %>%  
    group_by(Species) %>% 
    nest() %>% 
    mutate(plot = map2(data, Species, 
                       ~ggplot(data = .x, 
                               aes(x = Petal.Length, y = Petal.Width)) + 
                           geom_point() + ggtitle(.y)))
piris$plot[1]
piris$plot[3]
piris$plot[2]
```

# Working directory

Hadley Wickham on a [Stackoverflow
comment](https://stackoverflow.com/questions/4216753/check-existence-of-directory-and-create-if-doesnt-exist):

> "You should never use setwd() in R code - it basically defeats the idea of
> using a working directory because you can no longer easily move your code
> between computers. – hadley Nov 20 '10 at 23:44 "



# Further resources

## Blogs

* Gavin Simpson [My aversion to pipers](http://www.fromthebottomoftheheap.net/2015/06/03/my-aversion-to-pipes/) shows an Hadley tweet explaining that pipe might not be good in package development.

## Stackoverflow

* [Stack overflow question
  checklist](https://codeblog.jonskeet.uk/2012/11/24/stack-overflow-question-checklist/)
  general programming
* [How to make a great R reproducible
  example](https://stackoverflow.com/questions/5963269/how-to-make-a-great-r-reproducible-example) 


## Teaching R

* [Teach the tidyverse to beginners](http://varianceexplained.org/r/teach-tidyverse/)

* [Vectorization in R, Why?](http://www.noamross.net/blog/2014/4/16/vectorization-in-r--why.html)

* [Recommendations for resources for after workshops](https://github.com/carpentries/conversations/issues/18)

* [Rstudio education](https://education.rstudio.com/learn/)

