```{r developer_info, echo=FALSE}
# This template should be called by knit_reports.R 
#
# INPUT: data frames provided by knit_reports.R 
# OUTPUT: html report concerning a particular forest product
# Author: Paul Rougieux, European Forest Institute

```

```{r setup}
# See source for more details on input, output from the author
library(ggplot2)
library(reshape2)
library(plyr)
options(width=80)
dtf <- product.data$entity
agg <- product.data$eu_aggregates
trade <- product.data$trade
```
 
Explore `r product.data$metadata$title`
==========================
$$\beta_1$$
Consumption, production, trade and net trade in the European Union
-------------------------------------------------------------------
In the following file, we explore a data table containing `r product.data$metadata$title` consumption data for __`r length(unique(dtf$Country))`__ countries 
from `r min(dtf$Year)` to `r max(dtf$Year)`.
__In `r max(dtf$Year)`__, the overall EU consumption, production and trade 
in million `r product.data$metadata$unit` per item was:
```{r summarylastyear, echo=FALSE}
aggLast = subset(agg, Year==max(agg$Year)&
                       !Element%in%c("Import_Value", "Export_Value","Price"))
aggLast$Value = round(aggLast$Value/1e6)
aggLast = reshape(aggLast, drop=c("Year", "Deflator"),
                idvar="Item", timevar="Element",direction="wide")
# names(aggLast) = sub("Value.","",names(aggLast)) 
names(aggLast) = c("Item", "Consumption", "Production",
                     "Net_Trade", "Import", "Export")
print(aggLast) # This will be displaid on the HTML page
```

```{r consumptionTotalEU, fig.width=13, fig.height=6}
#summary(agg)
ggplot(data=subset(agg, !Element%in%c("Import_Value", "Export_Value","Price"))) +  
    geom_line(aes(x=Year, y=Value/1e6, colour=Item),size=1) +
    facet_wrap(~Element) +
    ylab(paste("Million",product.data$metadata$unit)) + theme_bw()
```


Consumption at the country level
--------------------------------
Apparent consumption of `r product.data$metadata$title` by country 
```{r consumptionbyCountry, fig.width=13, fig.height=8}
ggplot(data=dtf) +
    geom_line(aes(x=Year, y=Consumption/1e+06, colour=Item), size=1) + 
    facet_wrap(~Country, scales = "free_y") +
    xlab("Year") +
    ylab(paste("Million",product.data$metadata$unit)) + 
    theme_bw() +  theme(legend.position="bottom")
```


Prices
------
### EU prices
Average trade price of `r product.data$metadata$title` in the European Union
```{r priceEU, fig.width=10, fig.height=6}
# Rename Price to Price USD (Use level because Element is a factor)
levels(agg$Element)[levels(agg$Element)=="Price"] <- "Price_real_USD"
ggplot(data=subset(agg, Element=="Price_real_USD")) +
    geom_line(aes(x=Year, y=Value, color=Item)) +
    ylab(paste("Price in (2010) constant USD /", product.data$metadata$unit))
```

Comparison of nominal price in USD and real price in USD and EUR.
```{r priceComparison, fig.width=10, fig.height=12}
p <- dcast(agg, Year + Item ~ Element, value.var="Value")
# Add US and EUR deflator in the plot for information
p <- merge(p, EUR[c("Year", "DeflEUR")])
p <- merge(p, US[c("Year", "DeflUS")])
p <- mutate(p, 
            Price_Nominal_USD =  (Import_Value + Export_Value)/
                (Import_Quantity + Export_Quantity) *1000,
            Price_USD_2 = Price_EUR /ExchR)
p <- p[c("Year", "Item", "Price_Nominal_USD", "DeflUS", 
         "Price_real_USD", "DeflEUR", "ExchR", "Price_EUR")]
p <- melt(p, id=c("Year","Item"), variable.name="Element", value.name="Value")
ggplot(data=p) +
    geom_line(aes(x=Year, y=Value, color=Item)) +
        facet_wrap(~ Element, ncol=1, scales = "free_y")
```

### Trade prices by country
Prices expressed in constant US dollars of 2010 per `r product.data$metadata$unit`.
```
Import_Price = Import_Value / Import_Quantity / DeflUS*1000
Export_Price = Export_Value / Export_Quantity / DeflUS*1000
```
Trade prices for the 9 countries which have the highest trade volume in `r max(dtf$Year)` (volumes are in `r product.data$metadata$unit`).
```{r tradePrices9countries, echo=FALSE, warning=FALSE, results='asis'}
trade_last <- trade[grepl("Total", trade$Item) & trade$Year==max(dtf$Year),]
#order
trade_last <- trade_last[order(-trade_last$Quantity),]
# Remove duplicates (have to do this because germany is both a major exporter and importer)
# and select 9 highest trade volumes
countries <- unique(head(trade_last[!duplicated(trade_last$Country),]$Country, n=9))
trade_last <- reshape(subset(trade_last, Country%in%countries),
                      drop=c("Year", "FAOST_CODE", "Consumption", "GDPconstantUSD", "Item"),
                      idvar=c("Country", "Net_Trade"),
                      timevar="Trade",direction="wide")
print(xtable(arrange(trade_last, Net_Trade)), type = "html") 
```

```{r tradePricesByCountry, fig.width=12, fig.height=10, warning=FALSE, results='asis'}
ggplot(data=subset(trade, Country %in% countries)) +
    geom_line(aes(x=Year, y=Price_Trade, color=Item, linetype = Trade))+
    facet_wrap( ~ Country, ncol = 3) + 
    ylab(paste("Price in (2010) constant USD /", product.data$metadata$unit))
```


Plot log of the estimation data 
-------------------------------
We will estimate the model
$$ log(Consumption) = \beta_0 + \beta_1 log(GDP) + \beta_2 log(Price) +
\beta_3 log(Consumption_{t-1}) $$

Lets look at the relationship between log(Consumption) and log(GDP) first.
### Total `r product.data$metadata$title`
```{r, warning=FALSE}
plot(log(Consumption) ~ log(GDPconstantUSD),
     data=dtf[grep("Total", dtf$Item),])
points(log(Consumption) ~ log(GDPconstantUSD), 
     data=dtf[grepl("Total", dtf$Item) & dtf$Year>2011,], col="red")
```


### Explore the influence of Year and Net_Trade by country
Sort countries by Net_Trade, then display a color for each country
```{r warning=FALSE, fig.width=12, fig.height=8}
dtf_last = dtf[grepl("Total", dtf$Item) & dtf$Year==max(dtf$Year),]
dtf_last = dtf_last[order(dtf_last$Net_Trade),]
dtf$Country = factor(dtf$Country, levels= dtf_last$Country, ordered=TRUE)
p = ggplot(dtf, aes(x=log(GDPconstantUSD), y=log(Consumption))) + facet_wrap(~Item)
p + geom_point(aes(alpha=Year, color=Country))  
```


